Adaption of the Chumbley Score to matching of bullet striation marks

### 2018-01-03 doc_v2/doc_v2.Rmd
formatting

\tableofcontents

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "",
  fig.height = 6,
  fig.width = 6,
  fig.align = "center",
  out.width= '\\textwidth',
  cache = FALSE,
  fig.path='figures/',
  echo=FALSE,
  cache=TRUE
)
options(knitr.table.format = "latex")
library(tidyverse)
```

@@

## Problem statement and motivation

@@

\citep{chumbley} found that the algorithm gives false-positives in slightly over 2 cases (or 2%) of the time and about 9 false-negatives (or 9%) for every 100 comparisons. \citep{hadler} on the other hand, in the improved version of the algorithm first described by \citep{chumbley} found false-positives for 0 cases and 3 false-negatives (or 6%) for 50 comparisons of knwon matches.
\begin{figure}
\centering
\includegraphics[width=\textwidth]{images/B6-B2-L6.png}
```{r, echo=FALSE, fig.width=8, fig.height=3, out.width='\\textwidth'}
land <- read.csv("../data/b6b2l6.csv")
land %>% ggplot(aes(x = y, y=resid)) +geom_line() +theme_bw()
```

@@

```{r win-comparison, fig.width=8, fig.height=5, out.width='\\textwidth', fig.cap=" The two plots on the left show how the same shift behaves in case of a matching pair and the two plots on the right show how the different shift behaves in case of a matching pair."}
a<- matrix(NA, ncol = 3, nrow = 2)
toolmark.falsepositive<- data.frame(a)
colnames(toolmark.falsepositive)<- c("Classification", "Match", "Non-Match")
toolmark.falsepositive$Classification<- c("Match", "Non-Match")
toolmark.falsepositive$Match<- c(47,0)
toolmark.falsepositive$`Non-Match`<- c(3,50)

knitr::kable(toolmark.falsepositive, digits=0, format = "latex" ,booktabs=TRUE) %>% kableExtra::kable_styling(latex_options = c("hold"))

@@

## Potential limitations of the Chumbley Score

\hh{in the application to striation marks on bullet lands: smaller in width and curved - need to adjust parameter settings; something similar is done in \citet{afte-chumbley} for toolmark comparisons of slip-joint pliers. XXX work in }

Bullets are much smaller in length, width, are not flat and curved in the cross-sectional topography as opposed to tools like screw driver tips which produced longer and pronounced markings. This means the makings made by barrels in comparison to toolmarks may have a problem in distinctiveness. The majority of Bullet profiles and signatures extracted by procedures mentioned by  \citet{aoas} are almost 1/4 th the size of toolmarks as used by \citet{chumbley} or even smaller. Striations on Bullets are made on their curved surfaces, whereas the algorithm developed by \citet{chumbley} and \citet{hadler} has only been tested for flatter and wider surfaces which have negligible curvature. Therefore, using methods proposed for toolmarks may need adaptation in order to give tangible results for bullets. Moreover, in order to to get flat bullet signatures and remove the curvatures some kind of smoothing needs to be applied as a pre-step which needs further investigation as to whether the level of smoothing does effect the working of the algorithm on Bullets.

Also when in the optimization step, the Window of optimization for bullets will be shorter as the signatures are smaller. The idea is to keep the number of windows of optimization sufficiently large, which means shorter Trace segments (partition of signature or toolmark with length = size of window of optimization) that lets us compare smaller segments of one signature to another.

This introduces a problem as, if we go too small in the window of optimization, the unique features of the trace segments are lost and seem similar, while too large sizes vastly reduces the weight of small features that would otherwise uniquely classify a signature and hence identify the region of agreement.

Thus the Window of Optimization has a direct influence on whether we are
Falsely rejecting the null when it is true (Type I) or Falsely accepting the null when it is false (Type II) making identification of the optimum size of window of optimization very important.

@@

We used the adjusted Chumbley method as proposed in \citet{hadler} and implemented in the R package `toolmaRk` \citep{toolmark} on all pairwise land-to-land comparisons of the Hamby scans (a total of 85,491 comparisons) with the pairwise sets for the comparisons given in the following table.

```{r param, echo=FALSE, results='asis'}
param_setting<- all %>% select(wo, wv) %>% group_by(wo, wv) %>% tally() %>% select(wo,wv)
#param_setting<- data.frame(matrix(param_setting, nrow = 16))
#t(param_setting)
#param_setting %>% tally() 
library(kableExtra)
knitr::kable(t(param_setting)[,1:17], digits=0, caption = "Overview of parameter settings used for optimization and validation windows for bullet land signatures.", format = "latex" ,booktabs=TRUE) %>% kableExtra::kable_styling(latex_options = c("scale_down", "hold"))

knitr::kable(t(param_setting)[,18:33], digits=0, format = "latex" ,booktabs=TRUE) %>% kableExtra::kable_styling(latex_options = c("scale_down","hold"))

# library(xtable)
# strCaption <- "Overview of parameter settings used for optimization and validation windows for bullet land signatures."
#   
#   # set up xtable output
# print(xtable(t(param_setting)[,1:17], 
#              caption = strCaption, label = ""),
#       include.rownames = TRUE,
#       include.colnames = FALSE,
#       caption.placement = "top",
#       comment=FALSE,
#       floating=FALSE
#       )
# print(xtable(t(param_setting)[,17:33],label = ""),
#       include.rownames = FALSE,
#       include.colnames = FALSE,
#       caption.placement = "bottom",
#       comment=FALSE,
#       floating=TRUE
#       #table.placement = "!ht"
#       )
```

@@

The idea therefore is to first use these raw values of the profile directly in the chumbley algorithm, and see how the algorithm performs for different coarseness values (smoothing parameter as referred in the function lowess used in the chumbley algorithm)

@@

@@

@@

Figure \ref{fig:type2} gives an overview of type II error rates observed when varying the window size in the optimization step.

@@

Figure \ref{fig:type1} compares nominal (fixed) type I error and actually observed type I errors for the parameter settings in table \ref{tab:param}. With an increasing size of the window used in the optimization step the observed type I error rate decreases (slighty). This might be related to the increasing number of tests that fail for larger window sizes, in particular for non-matching striae (see fig \ref{fig:missings}).

```{r type1, fig.width=8, fig.height=5, out.width='\\textwidth', fig.cap="Comparison of observed and nominal type I error rates  across a range of window sizes for optimization $wo$. The horizontal line in each facet indicates the nominal type I error rate.  As the optimization window increase the observed type I error rate gets smaller. A smaller validation window tends to be associated with a higher type I error rate."}
errors %>% filter(wv %in% c(30,50)) %>%
  ggplot(aes(x = wo, y = actual, colour=factor(alpha))) + 
  geom_hline(aes(yintercept=alpha), colour="grey30") +
  geom_point(aes(shape=factor(wv))) +
  #  facet_grid(.~wo, labeller="label_both") + 
  geom_smooth(aes(group=alpha), se=FALSE, method="lm") +
  theme_bw() +
#  scale_y_log10(breaks=c(0.001,.005, 0.01, 0.05)) +
  scale_shape_discrete("Size of validation window, wv") +
  ylab("Observed type I error rate") +
  xlab(expression("Window size for optimization, "~wo)) +
  scale_colour_brewer(expression("Nominal type I error "~alpha), palette="Set2") +
  facet_wrap(~alpha, labeller="label_both", scales="free") +
  theme(legend.position="bottom")

>>>>

\tableofcontents
\newpage
```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "",
  fig.height = 6,
  fig.width = 6,
  fig.align = "center",
  out.width= '\\textwidth',
  cache = FALSE,
  fig.path='figures/',
  echo=FALSE,
  cache=TRUE
)
options(knitr.table.format = "latex")
library(tidyverse)
library(kableExtra)
```

@@

## Motivation

@@

\citep{chumbley} found that the algorithm gives false-positives in slightly over 2 cases (or 2%) of the time and about 9 false-negatives (or 9%) for every 100 comparisons. \citep{hadler} on the other hand, in the improved version of the algorithm first described by \citep{chumbley} found false-positives for 0 cases and 3 false-negatives (or 6%) for 50 comparisons of knwon matches.
```{r}
a<- matrix(NA, ncol = 3, nrow = 2)
toolmark.falsepositive<- data.frame(a)
colnames(toolmark.falsepositive)<- c("Classification", "Match", "Non-Match")
toolmark.falsepositive$Classification<- c("Match", "Non-Match")
toolmark.falsepositive$Match<- c(47,0)
toolmark.falsepositive$`Non-Match`<- c(3,50)

knitr::kable(toolmark.falsepositive, digits=0, format = "latex" ,booktabs=TRUE) %>% kableExtra::kable_styling(latex_options = c("hold"))
```

## Potential limitations of the Chumbley Score adaptation to bullets 
Bullets are much smaller in length, width, are not flat and curved in the cross-sectional topography as opposed to tools like screw driver tips which produced longer and pronounced markings. This means the makings made by barrels in comparison to toolmarks may have a problem in distinctiveness. The majority of Bullet profiles and signatures extracted by procedures mentioned by  \citet{aoas} are almost 1/4 th the size of toolmarks as used by \citet{chumbley} or even smaller. Striations on Bullets are made on their curved surfaces, whereas the algorithm developed by \citet{chumbley} and \citet{hadler} has only been tested for flatter and wider surfaces which have negligible curvature. Therefore, using methods proposed for toolmarks may need adaptation in order to give tangible results for bullets. Moreover, in order to to get flat bullet signatures and remove the curvatures some kind of smoothing needs to be applied as a pre-step which needs further investigation as to whether the level of smoothing does effect the working of the algorithm on Bullets.

Also when in the optimization step, the Window of optimization for bullets will be shorter as the signatures are smaller. The idea is to keep the number of windows of optimization sufficiently large, which means shorter Trace segments (partition of signature or toolmark with length = size of window of optimization) that lets us compare smaller segments of one signature to another. This introduces a problem as, if we go too small in the window of optimization, the unique features of the trace segments are lost and seem similar, while too large sizes vastly reduces the weight of small features that would otherwise uniquely classify a signature and hence identify the region of agreement.

Thus the Window of Optimization has a direct influence on whether we are Falsely rejecting the null when it is true (Type I) or Falsely accepting the null when it is false (Type II) making identification of the optimum size of window of optimization very important. This raises important questions about what parameter settings need to be chosen for bullet land comparison. Something similar is done in \citet{afte-chumbley} for toolmark comparisons of slip-joint pliers where optimum window sizes are determined. There is a need to figure out the best parameter settings which minimizes the errors for unsmoothed markings and pre-processed signatures as determined by \citet{aoas}. An analysis of these error rates and comparison with other methods will help us understand the adaptibility of the chumbley score to bullets.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{images/B6-B2-L6.png}
```{r, echo=FALSE, fig.width=8, fig.height=3, out.width='\\textwidth'}
land <- read.csv("../data/b6b2l6.csv")
land %>% ggplot(aes(x = y, y=resid)) +geom_line() +theme_bw()
```

@@

```{r win-comparison, fig.width=8, fig.height=5, out.width='\\textwidth', fig.cap=" The two plots on the left show how the same shift behaves in case of a matching pair and the two plots on the right show how the different shift behaves in case of a matching pair."}

@@

@@

@@

The idea therefore is to first use these raw values of the profile directly in the chumbley algorithm, and see how the algorithm performs for different coarseness values (smoothing parameter as referred in the function lowess used in the chumbley algorithm).

\pagebreak

@@

```{r param, echo=FALSE, results='asis'}
param_setting<- all %>% select(wo, wv) %>% group_by(wo, wv) %>% tally() %>% select(wo,wv)
#param_setting<- data.frame(matrix(param_setting, nrow = 16))
#t(param_setting)
#param_setting %>% tally() 
library(kableExtra)
#knitr::kable(t(param_setting)[,1:17], digits=0, caption = "Overview of parameter settings used for optimization and validation windows for bullet land signatures.", format = "latex" ,booktabs=TRUE) %>% kableExtra::kable_styling(latex_options = c("scale_down", "hold"))

par_tab1<-t(param_setting)[,1:18]
par_tab2<-t(param_setting)[,19:33]
knitr::kable(list(par_tab1,par_tab2), digits=0,caption = "Overview of parameter settings used for optimization and validation windows for bullet land signatures.", format = "latex" ,booktabs=TRUE) %>% kableExtra::kable_styling(latex_options = c("scale_down","hold"))%>% kableExtra::kable_styling(latex_options = c("scale_down","hold")) #%>% group_rows("",1,2) %>% group_rows("",3,4)

@@

We used the adjusted Chumbley method as proposed in \citet{hadler} and implemented in the R package `toolmaRk` \citep{toolmark} on all pairwise land-to-land comparisons of the Hamby scans (a total of 85,491 comparisons) with the pairwise sets for the comparisons given in the table \ref{tab:param}.

@@

Figure \ref{fig:type2} gives an overview of type II error rates observed when varying the window size in the optimization step. Two levels of validation window size 30 and 50 were chosen as to compare the error rates for different nominal type I errors. We notice that the trends for these nominal type I errors are similar and in most cases a validation window of 50 has higher type II error than for 30. A change in this trend is seen for a 0.05 $\alpha$ level, although the difference between the two windows is very small for this case. We can also notice an obvious trend of increase in the Type II error as the window of optimization increases and see a minimum around the optimization window size of 120 pixels. Hence we are inclined to choose a smaller validation window size and optimization window as 120.

@@

Figure \ref{fig:type1} compares nominal (fixed) type I error and actually observed type I errors for the parameter settings in table \ref{tab:param}. With an increasing size of the window used in the optimization step the observed type I error rate decreases (slighty)or As the optimization window increase the observed type I error rate gets smaller, while, A smaller validation window tends to be associated with a higher type I error rate. This can be better imagined for a given window of optimization, when the actual Type I error is comparable to the nominal level for only a select few validation window sizes and for comparable validation window sizes of 30 and 50 as done here, the actual type I error although increases negligibly, it does not seem to vary as much as it varies with the optimization window sizes. This might be related to the increasing number of tests that fail for larger window sizes, in particular for non-matching striae (see fig \ref{fig:missings}).

```{r type1, fig.width=8, fig.height=5, out.width='\\textwidth', fig.cap="Comparison of observed and nominal type I error rates  across a range of window sizes for optimization $wo$. The horizontal line in each facet indicates the nominal type I error rate."}
errors %>% filter(wv %in% c(30,50)) %>%
  ggplot(aes(x = wo, y = actual, colour=factor(alpha))) + 
  geom_hline(aes(yintercept=alpha), colour="grey30") +
  geom_point(aes(shape=factor(wv))) +
  #  facet_grid(.~wo, labeller="label_both") + 
  geom_smooth(aes(group=alpha), se=FALSE, method="lm") +
  theme_bw() +
#  scale_y_log10(breaks=c(0.001,.005, 0.01, 0.05)) +
  scale_shape_discrete("Size of validation window, wv") +
  ylab("Observed type I error rate") +
  xlab(expression("Window size for optimization, "~wo)) +
  scale_colour_brewer(expression("Nominal type I error "~alpha), palette="Set2") +
  facet_wrap(~alpha, labeller="label_both", scales="free") +
  theme(legend.position="bottom")
